Algebraic Groups\vspace{-1ex}

### 2019-09-27 Algebraic Geometry/Algebraic-Geometry.tex
Bump

%header stuff
\setlength{\headsep}{24pt}  % space between header and text
\pagestyle{fancy}     % set pagestyle for document
\lhead{Notes on the Schur-Weyl Functor} % put text in header (left side)
\rhead{Nico Courts} % put text in header (right side)
\cfoot{\itshape p. \thepage}
\setlength{\headheight}{15pt}
%\allowdisplaybreaks

@@

Fix a field $k$. 
\begin{defn}
	A \textbf{space with functions} is a topological spae $X$ along with a collection (a $k$-algebra!) $\calO(U)$ of maps $U\to k$ for each open 
	$U\subseteq X$.

@@

>>>>

%header stuff
\setlength{\headsep}{24pt}  % space between header and text
\pagestyle{fancy}     % set pagestyle for document
\lhead{Algebraic Geometry} % put text in header (left side)
\rhead{Notes by Nico Courts} % put text in header (right side)
\cfoot{\itshape p. \thepage}
\setlength{\headheight}{15pt}
%\allowdisplaybreaks

@@

Fix a field $k$. 
\begin{defn}
	A \textbf{space with functions} is a topological space $X$ along with a collection (a $k$-algebra!) $\calO(U)$ of maps $U\to k$ for each open 
	$U\subseteq X$.

@@

\section{September 27, 2019}
\begin{prob}
	Do all the exercises in Kempf chapter 1!
\end{prob}

For now we assume that $k$ is algebraically closed.

\subsection{Examples of spaces with functions}
There were lots of suggestions, but here are a couple:
\begin{ex}
	Let $X=\bbS^2$ and let $\calO_X^{cts}$ be the continuous $\bbC$-valued functions. Alternatively
	we could consider a different sheaf $\calO_X^{an}$, the holomorphic functions. Or we could 
	consider $\calO_X^\infty$, the $C^\infty$ functions (under some smooth structure).
\end{ex}

\begin{defn}
	A \textbf{morphism} of spaces with functions between $(X,\calO_X)$ and $(Y,\calO_Y)$ is a continuous map 
	$\varphi:X\to Y$ such that for all $U\subseteq Y$ open and $f\in\calO_Y(U)$, the function 
	\[\phi^\ast f=f\circ\phi|_{\phi^{-1}(U)}:\phi^{-1}(U)\to k\in \calO_X(\phi^{-1}(U))\]

	In other words, a morphism of spaces with functions is a map of spaces that \textit{respects the regular functions.}
\end{defn}
\begin{ex}
	Let $X,Y$ be topological spaces and let $\calO_X$ and $\calO_Y$ be the continuous functions. Then morphisms are just continuous maps.
\end{ex}
\begin{ex}
	When $X$ and $Y$ are manifolds and $\calO_\bullet$ are complex-valued functions, then the maorphisms are maps 
	of manifolds.
\end{ex}

So now we return to the examples we saw before: $(\bbS^2,\calO^\infty)$, $(\bbS^2,\calO^{cts})$, and $(\bbS^2,\calO^{an})$.
A natural question to ask is when we have morphisms between these spaces to see if there exist ones that are the identity on $\bbS^2$.

Consider the identity map from the continuous to the analytic functions. Then take any map $f\in\calO^{an}$ and consider that 
\[f=f\circ id_{id^{-1}(U)}:U\to k\in\calO^{cts}(U)\]
and there is no map in the other direction.
\begin{rmk}
	Notice that since we are pulling functions back, the maps go in the opposite direction as you may think at first.
\end{rmk}

We can also talk about \textbf{open subspaces}. If $V\subseteq X$ is an open subset, we can let the induced space with functions 
be $(V,\calO_V)$ where if $U\subseteq V$ then $\calO_V(U):=\calO_X(U)$.

\subsection{Varieties}
\begin{defn}
	An \textbf{affine $k$-variety} is a space with functions $(Y,\calO_Y)$ such that for every space with functions $(X,\calO_X)$,
	the natural map 
	\[\Hom((X,\calO_X),(Y,\calO_Y))\to\Hom_{\Alg}(\calO_Y(Y),\calO_X(X))\]
	is a bijection and furthermore $\calO_Y(Y)=:k[Y]$ is a finitely generated $k$-algebra.
\end{defn}
\begin{rmk}
	The idea here is that the algebra maps (on the right) are precisely the same as the geometry maps (on the left).
	Algebraic geometry, baby.
\end{rmk}
So then this leads to a very simple (loose) definition:
\begin{defn}
	A \textbf{variety} is something that is covered by affine varieties.
\end{defn}

\begin{ex}
	$\bbA^1=k$. Give this space the cofinite topology. Then if we have $U=k\setminus\{x_1,\dots,x_n\}\subset \bbA^1$,
	\[\calO_{\bbA^1}(U)=\{f(t)\in k(t)| \text{poles are in }\{x_i\}\}\]
\end{ex}
\begin{prob}
	Show that $\bbA^1$ is an affine variety!
\end{prob}
\begin{rmk}
	Notice that this staetment is equivalent to saying that any morphism of spaces with functions gives us a regular map $X\to k$.
\end{rmk}

### 2019-09-27 Algebraic Groups/Algebraic-Groups.tex
Bump

% Document-Specific Macros
\DeclareMathOperator{\1}{\mathbbm{1}}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\Sym}{Sym}

\begin{document}
%make the title page
\title{Algebraic Groups\vspace{-1ex}}
\author{A course by Jarrod Alper and Julia Pevtsova\\
Notes by Nico Courts}
\date{Autumn 2019/ Winter and Spring 2020}
\maketitle

@@

\subsection{Some Interesting Facts}
If we had to write down five results that we'd like to get out of this class:
\begin{prop}
	Every affine group scheme of finite type over a field embeds into $GL_n$ as a closed subgroup.
\end{prop}
\begin{thm}[Cheballey's Theorem]
	Let $G$ be a finite type group scheme over a field. Then it factors as 
	\[1\to H\to G\to A\to 1\]
	where $A$ is abelian and $H$ is affine (linear algebraic).
\end{thm}
\begin{prop}
	If $G$ is an affine group scheme of finite type over $k$, then we have af actorization
	\[1\to U\to G\to R\to 1\]
	where $U$ is unipotent and $R$ is reductive.
\end{prop}
\begin{prop}
	$H\subseteq G$  a subgroup scheme. Then $G/H$ is a projective scheme.
\end{prop}
Finally we want to talk about Tanakka duality and how the representations of $G$ define $G$ itself.

\end{document}

>>>>

% Document-Specific Macros
\DeclareMathOperator{\1}{\mathbbm{1}}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\Sym}{Sym}
\DeclareMathOperator{\Spec}{Spec}

\begin{document}
%make the title page
\title{Algebraic Groups\vspace{-1ex}}
\author{A course by Jarod Alper and Julia Pevtsova\\
Notes by Nico Courts}
\date{Autumn 2019/ Winter and Spring 2020}
\maketitle

@@

\subsection{Some Interesting Facts}
If we had to write down five results that we'd like to get out of this class:
\begin{prop}
	Every affine group scheme of finite type over a field embeds into $GL_n$ as a closed subgroup.
\end{prop}
\begin{thm}[Chevalley's Theorem]
	Let $G$ be a finite type group scheme over a field. Then it factors as 
	\[1\to H\to G\to A\to 1\]
	where $A$ is abelian and $H$ is affine (linear algebraic).
\end{thm}
\begin{prop}
	If $G$ is an affine group scheme of finite type over $k$, then we have af actorization
	\[1\to U\to G\to R\to 1\]
	where $U$ is unipotent and $R$ is reductive.
\end{prop}
\begin{prop}
	$H\subseteq G$  a subgroup scheme. Then $G/H$ is a projective scheme.
\end{prop}
Finally we want to talk about Tanakka duality and how the representations of $G$ define $G$ itself.

\section{September 27th, 2019}
Last time we defined a group scheme (a group object in the category of schemes over a base scheme). We also mentioned that 
You could define it as a map $h_G:\mathbf{Sch}/S\to \mathbf{Set}$ along with a factorization through $\mathbf{Grp}$.

We defined an \textbf{algebraic group} over $k$ as a group scheme over $\operatorname{Spec} k$ of finite type and a \textbf{linear algebraic group}
to be an \textit{affine} group scheme over $k$ of finite type.

\subsection{Hopf Algebras}
Let $G=\Spec A$ be a linear algebraic group over $k$. I have seen most of these before (see Waterhouse or my Hopf algebra notes)
\begin{rmk}
	One think I haven't seen explicitly before: Notice that the augmentation ideal $\ker \varepsilon$, where $\varepsilon$ is the counit,
	is the (maximal!) ideal corresponding in the algebro-geometric sense to the identity element in $G$.
\end{rmk}
\begin{defn}
	A \textbf{Hopf algebra} is ...
\end{defn}
\begin{defn}
	Let $G$ be an algebraic group over $k$. Then if $h_G$ factors through $\Ab$, $G$ is called \textbf{commutative.}
\end{defn}

\subsection{Some Examples}
\begin{rmk}
	Note that to define a functor from schemes over $k$, is suffices to define it on affine schemes, thereby defining 
	the (Zariski) local behavior of any such map. Thus we really only need to consider maps in $\Alg$.
\end{rmk}
\begin{itemize}
	\item $\bbG_a$. Here we can define it as a functor that sends $S\mapsto\Gamma(S,\calO_S)$. Geometrically, $\bbG_a=\bbA^1$ where the multiplication is addition, inverses send $x\mapsto -x$ and the unit is the zero map.
	The Hopf algebraic picture is the usual dual thing.
	\item $\bbG_m$ as a scheme isthe map $S\mapsto \Gamma(S,\calO_S)^\ast$. In the geometric picture, $\bbA^1\setminus\{0\}$ and the algebra structure comes from multiplciation. Hopf is pretty easy.
	\item $\GL_n$ is a scheme that sends
	\[S\mapsto \left\{A=(a_{ij}): a_{ij}\in\Gamma(S,\calO_S), \det(A)\in\Gamma(S,\calO_S)^\ast\right\}\]
	the algebra is $\bbA^{n\times n}\setminus \{\det = 0\}$ with the usual multiplication. The coalgebra structure can be seen in the book.
\end{itemize}
This one requires some more explaination so I am setting it apart.
\begin{ex}
	Let $V$ be a finite dimensionatl vector space over $k$. Then we can define the algebraic group $V_a$ which sends 
	\[S\mapsto \Gamma(S,\calO_S)\otimes_k V.\]
	Geometrically we are looking at $\bbA(V)=\Spec\Sym ^\ast V^\vee\simeq \Spec k[x_1,\dots,x_n]$ where $n=\dim V$.
\end{ex}

What about finite groups? As a scheme, we want $G=\sqcup_{g\in G}\Spec k$. The functor sends $S\mapsto \operatorname{Mor}_{\operatorname{Set}}(\pi_0(S),G)$,
or maps from the connected components into $G$.

\begin{ex}
	Now consider the $n^{th}$ roots of unity: as a scheme, $\mu_n=\Spec k[t]/(t^n-1)\subseteq \bbG_m$.
	If both $k=\bar k$ and $\ch k\nmid n$, then $\mu_n\cong\bbZ/n\bbZ$.

	But if (e.g.) $k=\bbQ$, then $\mu_3$ is $\bbQ[t]/(t^3-1)=\Spec\bbQ\sqcup \Spec\bbQ(\xi)$ where $\xi$ is a primitive third root of unity.

	If, on the other hand, $k=\bar \bbF_3$ and consider $\mu_3$, we get a single point with residye field $\bar\bbF_3$.
\end{ex}

\begin{ex}
	If we are in the case of positive characteristic, then we get an algebraic group $\alpha_p$. Here the scheme is $\Spec k[x]/x^p$ and functorially it 
	is the map $S\mapsto \{F\in\Gamma(S,\calO_S)|f^p=0\}$.
\end{ex}

\subsection{Matrix Groups}
We already defined $\GL_n$, but we can also define 
\[\operatorname{SL}_n:S\mapsto\{A=(a_{ij})|\det A=1\}\]
with scheme $\Spec k[x_{ij}]/(\det-1)$.

We also have the (upper) triangular matrices $T_n$ and unitary group $U_n$ and diagonal group $D_n$

\begin{defn}
	Let $G$ be a linear algebraic group. Then 
	\begin{itemize}
		\item $G$ is a \textbf{vector group} if $G\cong V_a$ for some finite dimensional $V$.
		\item $G$ is a \textbf{split torus} if $G\cong \bbG_m^n$.
		\item $G$ is a \textbf{torus} if there is a field extention $k\to k'$ such that 
		\[G\times_{\Spec k}\Spec k'\cong \bbG^n_{m,k'}\]
	\end{itemize}
\end{defn}
\end{document}
